\textit{k}-mer-based GWAS enhances the discovery of causal variants and candidate genes in soybean

### 2023-07-04 additional_files/manuscript.tex
Address most of the reviewers' comments

Genome-wide association studies (GWAS) are powerful statistical methods that
detect associations between genotype and phenotype at genome scale. Despite
their power, GWAS frequently fail to pinpoint the causal variant or the gene
controlling a given trait in crop species. Assessing genetic variants other than
single-nucleotide polymorphisms (SNPs) could alleviate this problem.  In this
study, we tested the potential of SV- and \emph{k}-mer-based GWAS in soybean by
applying these methods as well as conventional SNP/indel-based GWAS to 13
traits. We assessed the performance of each GWAS approach based on loci for
which the causal genes or variants were known from previous genetic studies. We
found that \emph{k}-mer-based GWAS was the most versatile approach and the best
at pinpointing causal variants or candidate genes.  Moreover,
\textit{k}-mer-based analyses identified promising candidate genes for loci
related to pod color, pubescence form, and resistance to \textit{Phytophthora
sojae}. In our dataset, SV-based GWAS did not add value compared to
\textit{k}-mer-based GWAS and may not be worth the time and computational
resources invested.  Despite promising results, significant challenges remain
regarding the downstream analysis of \emph{k}-mer-based GWAS.  Notably, better
methods are needed to associate significant \emph{k}-mers with sequence
variation. Our results suggest that coupling \emph{k}-mer- and SNP/indel-based
GWAS is a powerful approach for discovering candidate genes in crop species.

@@

We selected 389 inbred \emph{G. max} accessions based on the availability of
Illumina WGS data and phenotypic data for resistance to the oomycete \emph{P.
sojae} \citep{deronne2022}.  We identified 741 SRA runs corresponding to
Illumina paired-end sequencing data for those 389 accessions \citep[data from][]{zhou2015, valliyodan2016, fang2017, bayer2021}
and extracted paired-end reads using the fastq-dump command v. 2.9.6
(\citeauthor{sratoolkit}) with the
option \texttt{-\-split-3}.
Reads that were not paired in the raw sequencing data were not used
for downstream analyses.  The runs downloaded from the SRA and associated
metadata are listed in Supplemental file 2.

We first filtered the sequencing data for quality and the presence of sequencing
adapters using bbduk from BBtools v. 38.25
(\citeauthor{bbtools}). Reads
were then mapped to the reference assembly version 4 of \emph{G. max}
cultivar Williams82 \citep{valliyodan2019} concatenated with chloroplast
and mitochondrion sequences obtained from SoyBase \citep{grant2010}. We used
bwa-mem v. 0.7.17-r1188 \citep{li2009-bwa} with default parameters to map the
reads in paired-end mode. Reads that were left unpaired following adapter- and
quality-trimming were aligned separately in single-end mode with bwa-mem and
merged with mapped paired-end reads using samtools \texttt{merge}
v. 1.12 \citep{li2009-samtools}.  We then added read groups identifying
individual SRA runs to the BAM files and merged the reads belonging to the same
accession using bamaddrg
(\citeauthor{bamaddrg}). We used the
resulting BAM files for all downstream analyses requiring aligned reads.
Metadata on mapping depth of aligned data can be found in Supplemental file 3.

@@

Moreover, we pruned the dataset using the \texttt{-\-indep-pairwise} option of
PLINK v. 1.90b5.3 \citep{purcell2007} with a window of 1,000 markers, a
step of 100 markers and an $r^2$ LD threshold of 0.9. This pruning step was
implemented to reduce the computational requirements of the GWAS analysis by
removing co-segregating markers. Markers located on unanchored scaffolds were
also removed. The resulting dataset was used for GWAS and comprised 773,060
SNPs and 151,570 indels.

@@

We used the approach outlined by \cite{voichek2020} for \emph{k}-mer-based GWAS
and their recommendations on the
associated GitHub page (v. 0.2-beta, \url{https://github.com/voichek/kmersGWAS}) to generate
a presence/absence \emph{k}-mer table for use in GWAS.  The first step of
the approach involved counting all \emph{k}-mers of length 31 (31-mers) present
in the trimmed (FASTQ) reads for each accession using KMC3 v. 3.2.1
\citep{kokot2017}. \emph{k}-mers were counted twice, the first time using their
canonized form (the first of either the observed \textit{k}-mer or its reverse
complement in lexicographical order) and the second time using the observed
\emph{k}-mers themselves (non-canonized form). The two sets of \textit{k}-mers
were then combined and filtered by
keeping only those seen in at least 5 accessions, and observed in both canonized
and non-canonized form in at least 20\% of the accessions in which they are found.
We then generated a table indicating the presence or absence of each
\emph{k}-mer in all accessions. This table was used to compute a kinship
table for use in GWAS and was also used directly as input genotypes for the
GWAS analyses. We used \emph{k}-mers with a MAF $\geq$ 0.02 for both the
computation of kinship and GWAS.

@@

We used the GAPIT3 R package v. 3.1.0 \citep{wang2021} for conducting GWAS on
the SNP/indel and SV datasets described above. SVs and indels required some additional
preprocessing steps to make variant
representation suitable for downstream analyses \citep{lemay2022gwas}. Briefly, we recoded all
variants as SNPs prior to converting the VCF files to Hapmap diploid format
using the TASSEL command-line tools \citep{bradbury2007}; variant IDs enabled
retrieval of the proper metadata for each variant after GWAS.

We ran GAPIT using a mixed linear model (MLM) with 9 principal components and the VanRaden
algorithm for computing the kinship matrix. We used an MLM model for
consistency with the statistical model used by the \textit{k}-mer-based
approach.  We used a randomization approach to determine the 5\% family-wise
error-rate threshold for each GWAS analysis as described by \cite{voichek2020}.
To compute this threshold, we permuted the phenotypic observations 100 times
and computed a GWAS on these permuted phenotypes to obtain a distribution of
top \emph{p}-values under the null hypothesis. The fifth most significant of
the top 100 \emph{p}-values was used as the significance threshold for
inference, i.e. markers with \emph{p}-values lower than this threshold were
considered significant.

We conducted GWAS based on \emph{k}-mers using the method developed by
\cite{voichek2020}. In brief, the
analysis was conducted in two steps. The first step used an approximate model
to identify 1 million potentially significant \emph{k}-mers. These were then used in
the second step as input to an exact model implemented in GEMMA \citep{zhou2012}, resulting
in a \emph{p}-value for each \emph{k}-mer. We
used the kinship computed from the \emph{k}-mers to correct for relatedness
between accessions. The program automatically outputs a list of \emph{k}-mers
that passed a 5\% family-wise error-rate threshold determined by a
randomization approach, which we considered significant for the purposes of
downstream analyses.

@@

\emph{k}-mers are not intrinsically associated with a particular genomic region
and a tailored analysis was therefore required to associate them with genomic
coordinates. To do so, we queried the mapped reads (in BAM format) of all accessions for matches
to significant \emph{k}-mers for a given trait. Because this is a
computationally demanding operation (there were typically tens of thousands of
significant \emph{k}-mers per trait), we developed a C program called katcher
(\url{https://github.com/malemay/katcher}) that
uses the htslib library \citep{bonfield2021} for efficient reading and
writing of BAM files. katcher and its associated utilities allow for efficient
retrieval and annotation of mapped reads containing any \emph{k}-mer from a set
of interest.

Once we retrieved all the reads containing significant \emph{k}-mers, we next
needed to gather the mapping information contained in those reads for the
purposes of identifying contiguous signals and comparing them to annotations. To
simplify the amount of information to process in downstream analyses, we
considered only the most significant \emph{k}-mer contained in a given read and
the position where this \emph{k}-mer was most often observed in
the dataset. Moreover, we limited the analysis to reads with a minimum mapping
quality of 20 and to \emph{k}-mers that were observed at least 10 times in at
least one accession. The latter was done to remove spurious \emph{k}-mers that may originate
from sequencing errors. All downstream analyses (Manhattan plots,
identification of signals, analysis of genes located near significant
\emph{k}-mers, etc.) used this processed set of significant \emph{k}-mers.

@@

\cite{zabala2003} identified Glyma.06g202300 as the gene associated with the
\textit{T} locus for pubescence color and found a 1-bp deletion as a putative
causal variant at this locus. For all three GWAS approaches, we found a signal
overlapping this gene (Supplemental Figure S6). In
particular, the regions defined by the 1\% most associated variants
identified by the SNP- and \textit{k}-mer-based approaches included
Glyma.06g202300 (Supplemental Figure S7). While the most
significant \textit{k}-mers did occur within the sequence of this gene, they
did not correspond to the documented causal variant (Supplemental Figure
S8d).  Still, the fourth and fifth most strongly
associated \textit{k}-mers corresponded to this causal variant (Supplemental Figures
S8d and S9). We
observed that not all accessions with gray pubescence harbored the deleted
nucleotide (Supplemental Figure S9c), which could indicate
either that other causal variants are resulting in a non-functional F3’H in
this population, or that the causal variant lies elsewhere. As for SNP-based GWAS,
the most significant variant within the sequence of Glyma.06g202300 was the
causal variant, however it was not the most significantly associated variant
overall (Supplemental Figures S7b and
S8b). SV-based GWAS did not detect the causal variant
at this locus, as expected from its size.

@@

% Pubescence color - Td locus
We conducted a separate GWAS analysis contrasting only accessions with tawny
and light tawny pubescence because the \textit{Td} locus is known to control
these differences in color. Accordingly, we detected signals in this region
using all three approaches, but only SNP- and \textit{k}-mer-based GWAS
identified a signal that overlapped Glyma.03g258700 (Figure
\ref{pubescence-color-main-figure}a-b, and Supplemental Figures S10 and
S11).  While the most significant
\textit{k}-mer identified the causal variant documented at this locus
\citep{yan2020}, it was the fourth most significantly associated variant for
the SNP-based approach (Figure \ref{pubescence-color-main-figure}b-d, Supplemental Figure 
S12).  Interestingly, some of the accessions
bearing the haplotype associated with tawny pubescence displayed light tawny
pubescence, suggesting that there may be more than one causal variant at this
locus (Figure \ref{pubescence-color-main-figure}e).

@@

% Seed coat color - I locus
The causal variant at the \textit{I} locus for seed coat color is a complex SV
that takes the form of a 10.91-kb inverted duplication of three chalcone
synthase (CHS) genes \citep[CHS1, CHS3 and CHS4;][]{tuteja2008}. The presence of
this inverted duplication results in the silencing of all CHS genes in the seed
coat specifically and thus in the absence of seed coat pigmentation
\citep{tuteja2009}. While we did find signals overlapping this causal variant
using all three GWAS approaches (Supplemental Figures S13 and
S14), none of the approaches identified
it. This failure to detect the causal variant is likely due
to its complexity, which makes it difficult to detect and genotype it. While it
is likely that at least some of the significantly associated \textit{k}-mers
actually are derived from this causal variant, the lack of systematic methods for
linking \textit{k}-mers to sequence variation hindered our ability to do so.

@@

While all GWAS approaches found signals near the Glyma.19g194300 gene
associated with the \textit{Dt1} locus for stem termination type, the SNP- and
\textit{k}-mer-based analyses performed best since the region defined by their
most highly associated markers (top 1\%) included the causal gene (Supplemental Figures
S18 and S19).
We did identify one of the SNPs suggested as putative causal variants for this
locus by \cite{tian2010} and \cite{liu2010}, however this SNP was only the
135\textsuperscript{th} most associated marker found by Platypus and the
12\textsuperscript{th} most associated \emph{k}-mer (Supplemental Figures
S20 and S21).
Our results are consistent with previous findings in suggesting that there is
more than one causal variant at the \textit{Dt1} locus \citep{liu2010,
tian2010}. Indeed, a few cultivars bearing the allele associated with
\emph{dt1} showed indeterminate or semi-determinate phenotypes and some
cultivars showed a determinate phenotype without bearing the allele associated
with \emph{dt1} (Supplemental Figure S21c). The existence of
several causal variants might explain why our analyses did not identify a
causal variant more clearly.

@@

We conducted a GWAS analysis that only compared accessions with black and brown
hila in an attempt to detect stronger signals at the \textit{R} locus.  We
accordingly detected a signal overlapping this locus using all approaches, but
only SNP- and \textit{k}-mer-based GWAS detected signals overlapping the
Glyma.09g235100 gene associated with \textit{R} (Supplemental Figures
S22 and S23).
\cite{gillman2011} identified this R2R3 MYB transcription factor as the
molecular basis for the \textit{R} locus and documented four different
loss-of-function alleles linked to brown hilum color. Although our analyses did
detect strong signals in the vicinity of this gene, the most associated markers
were not located in the body of the gene (Supplemental Figure
S23).  Interestingly, the most significantly
associated \emph{k}-mers detected by our analysis mapped 2 kb upstream of
another gene (Glyma.09g234900) annotated as a MYB transcription factor and
putatively involved in the regulation of anthocyanin biosynthesis. This gene as
well as Glyma.09g235100 and two others are part of a tandem array of four R2R3
MYB genes considered as candidates by \cite{gillman2011}, and of which
Glyma.09g235100 was the only one expressed in the seed coat. Given the
convincing evidence provided by \cite{gillman2011}, it is most likely that
Glyma.09g235100 is indeed the gene associated with the \emph{R} locus and that
our failure to identify the most significantly associated \emph{k}-mers as
causal variants is due to the coexistence of several variants that can cause
loss of function at the \emph{R} locus.  Still, we observed two causal variants
previously documented by \cite{gillman2011} in our dataset (Supplemental Figures
S24 and S25).

@@

We conducted a GWAS analysis comparing accessions with normal and semi-sparse
pubes\-cence in order to detect loci associated with pubescence density. This analysis
detected signals at the \textit{Ps} locus using all three approaches (Figure
\ref{pubescence-density-main-figure}a-b and Supplemental Figures S26, S27
and S28). The causal
variant at this recently cloned locus is a copy number variant (CNV) overlapping the
Glyma.12g187200 gene, with higher copy number resulting in reduced pubescence \citep{liu2020ps}.
We identified a region of $\sim$42 kb that varied in copy number across
the population and was associated with the trait (Figure
\ref{pubescence-density-main-figure}c), and therefore possibly corresponded to the
25.6-kb region identified as the causal CNV by \cite{liu2020ps}. We found that
reads containing the most significantly associated \textit{k}-mer mapped to either of the two ends
of the CNV (Figure \ref{pubescence-density-main-figure}d). This suggests that
the most associated \textit{k}-mer is indeed linked to the CNV
that is causal at this locus, while \textit{k}-mers within the
CNV region may represent differences in sequence between the copies.

@@

The difference in size between the CNV reported by \cite{liu2020ps} and our
results may be due to the presence of some unresolved sequence (a stretch of
4,635 N nucleotides) over that interval in the reference sequence that we used.
Because our SV filtering process removed any alternate allele containing N nucleotides,
we reasoned that the CNV may have been found by our pipeline but
filtered out later on. Indeed, we found a 42.8-kb duplication corresponding
to the one described above among the genotype calls made by smoove. An association
analysis of that single variant using GAPIT found a \textit{p}-value of 1.9e-23,
which would have made it the most significant SV. In this particular case, the
failure of the SV-based GWAS analysis to find identify the causal variant was due
to our filtering parameters and not to limitations of the SV genotype calls per se.

@@

We conducted a GWAS analysis comparing dull and shiny seed coat phenotypes for
the seed coat luster trait. Seed coat luster has been shown by
\cite{gijzen1999} to be largely caused by the deposition of a hydrophobic
protein (HPS) at the seed surface. Copy number variation of this gene at the
\textit{B} locus was later shown to explain variation in luster, although
sequence variation in the HPS sequence may also play a role \citep{gijzen2006}.
Using a sequencing depth-based analysis, we identified a 31-kb region spanning
positions 9,386,109-9,417,431 on chromosome Gm15 that exhibited variation in
copy number across accessions. This CNV overlapped four genes annotated as
containing HPS domains and therefore likely represented the causal variant
(Supplemental Figures S29 and
S30).  We found signals overlapping this
CNV using all approaches, and the most significantly associated \textit{k}-mer
was notably located within the CNV region (Supplemental Figures
S29 and
S31). However, we were not able to
directly link any of the variants or \textit{k}-mers identified to the causal
CNV.

@@

Both the \textit{L1} and the \textit{L2} loci for pod color have yet to be
cloned. While \cite{he2015} proposed the gene Glyma.19g101700 as likely
corresponding to \textit{L1} based on fine mapping and expression analysis, our
results as well as those reported by \cite{bandillo2017} do not support this
gene as a candidate due to its location $\sim$3 Mb away from GWAS signals. Our
analyses based on \textit{k}-mers and SNPs found signals overlapping 14 and
6 genes, respectively (Supplemental figures S32 and
S33). Of these, a MATE transporter (Glyma.19g120300)
appears as a prime candidate as the most significant \emph{k}-mer was observed
within its sequence and it is highly expressed in pods according to data hosted
on SoyBase \citep{grant2010, severin2010}. Moreover, the role of some MATE
proteins in transporting flavonoids has been demonstrated \citep{chen2015}.

@@

In addition to \emph{k}-mers, we also assessed the use of SVs in GWAS analysis
of soybean.  Although we did find the causal variant as the most significantly
associated SV at the \emph{W1} locus, results were otherwise rather
inconclusive.  Indeed, most of the other known or putative causal variants were
SNPs or indels, and therefore were not represented in the SV datasets. At most
loci where SVs are known to be the underlying causal variant (e.g.  \emph{I},
\emph{B}, \emph{Ps}), the analyses based on SVs did not find the causal
variants, most likely due to the difficulty of calling and genotyping such
variants \citep{kosugi2019, lemay2022}. In the case of the \textit{Ps} locus,
this limitation was due to a filtering step in our pipeline rather than to an
inherent inability to properly call and genotype the causal variant.

>>>>

Genome-wide association studies (GWAS) are powerful statistical methods that
detect associations between genotype and phenotype at genome scale. Despite
their power, GWAS frequently fail to pinpoint the causal variant or the gene
controlling a given trait in crop species. Assessing genetic variants other than
single-nucleotide polymorphisms (SNPs) could alleviate this problem.  In this
study, we tested the potential of structural variant (SV)- and \emph{k}-mer-based GWAS in soybean by
applying these methods as well as conventional SNP/indel-based GWAS to 13
traits. We assessed the performance of each GWAS approach based on loci for
which the causal genes or variants were known from previous genetic studies. We
found that \emph{k}-mer-based GWAS was the most versatile approach and the best
at pinpointing causal variants or candidate genes.  Moreover,
\textit{k}-mer-based analyses identified promising candidate genes for loci
related to pod color, pubescence form, and resistance to \textit{Phytophthora
sojae}. In our dataset, SV-based GWAS did not add value compared to
\textit{k}-mer-based GWAS and may not be worth the time and computational
resources invested.  Despite promising results, significant challenges remain
regarding the downstream analysis of \emph{k}-mer-based GWAS.  Notably, better
methods are needed to associate significant \emph{k}-mers with sequence
variation. Our results suggest that coupling \emph{k}-mer- and SNP/indel-based
GWAS is a powerful approach for discovering candidate genes in crop species.

@@

We selected 389 inbred \emph{G. max} accessions based on the availability of
phenotypic data for resistance to the oomycete \emph{P.
sojae} \shortcite{deronne2022} and matching Illumina WGS data. We identified 741 SRA runs corresponding to
Illumina paired-end sequencing data for those 389 accessions \shortcite<data from>[]{zhou2015, valliyodan2016, fang2017, bayer2021}
and extracted paired-end reads using the fastq-dump command v. 2.9.6
(\citeauthor{sratoolkit}) with the
option \texttt{-\-split-3}.
Reads that were not paired in the raw sequencing data were not used
for downstream analyses.  The runs downloaded from the SRA and associated
metadata are listed in Supplemental file 2.

We first filtered the sequencing data for quality and the presence of sequencing
adapters using bbduk from BBtools v. 38.25
(\citeauthor{bbtools}). Reads
were then mapped to the reference assembly version 4 of \emph{G. max}
cultivar Williams82 \shortcite{valliyodan2019} concatenated with chloroplast
and mitochondrion sequences obtained from SoyBase \shortcite{grant2010}. We used
bwa-mem v. 0.7.17-r1188 \shortcite{li2009-bwa} with default parameters to map the
reads in paired-end mode. Reads that were left unpaired following adapter- and
quality-trimming were aligned separately in single-end mode with bwa-mem and
merged with mapped paired-end reads using samtools \texttt{merge}
v. 1.12 \shortcite{li2009-samtools}.  We then added read groups identifying
individual SRA runs to the BAM files and merged the reads belonging to the same
accession using bamaddrg
(\citeauthor{bamaddrg}). We used the
resulting BAM files for all downstream analyses requiring aligned reads.
Metadata on mapping depth of aligned data can be found in Supplemental file 3.

@@

Moreover, we pruned the dataset using the \texttt{-\-indep-pairwise} option of
PLINK v. 1.90b5.3 \shortcite{purcell2007} with a window of 1,000 markers, a
step of 100 markers and an $r^2$ LD threshold of 0.9. This pruning step was
implemented to reduce the computational requirements of the GWAS analysis by
removing co-segregating markers. Markers located on unanchored scaffolds were
also removed. The resulting dataset was used for GWAS and comprised 773,060
SNPs and 151,570 indels. Pruned markers were still used for GWAS at targeted
loci when filtered markers identified significant associations (more details
in the section describing the analysis of significant signals).

@@

We used the approach outlined by \shortciteA{voichek2020} for \emph{k}-mer-based GWAS
and their recommendations on the
associated GitHub page (v. 0.2-beta, \url{https://github.com/voichek/kmersGWAS}) to generate
a presence/absence \emph{k}-mer table for use in GWAS.  The first step of
the approach involved counting all \emph{k}-mers of length 31 (31-mers) present
in the trimmed (FASTQ) reads for each accession using KMC3 v. 3.2.1
\shortcite{kokot2017}. We selected $k = 31$ based on demonstrated performance in previously
published studies \shortcite{rahman2018, voichek2020, he2021}.
\emph{k}-mers were counted twice, the first time using their
canonized form (the first of either the observed \textit{k}-mer or its reverse
complement in lexicographical order) and the second time using the observed
\emph{k}-mers themselves (non-canonized form). The two sets of \textit{k}-mers
were then combined and filtered by
keeping only those seen in at least 5 accessions, and observed in both canonized
and non-canonized form in at least 20\% of the accessions in which they are found.
We then generated a table indicating the presence or absence of each
\emph{k}-mer in all accessions. This table was used to compute a kinship
table for use in GWAS and was also used directly as input genotypes for the
GWAS analyses. We used all \emph{k}-mers with a MAF $\geq$ 0.02 as input genotypes
for both the computation of kinship and GWAS analysis. The total number of \textit{k}-mers
in this filtered table was 637,605,529.

@@

We used the GAPIT3 R package v. 3.1.0 \shortcite{wang2021} for conducting GWAS on
the SNP/indel and SV datasets described above. SVs and indels required some additional
preprocessing steps to make variant
representation suitable for downstream analyses \shortcite{lemay2022gwas}. Briefly, we recoded all
variants as SNPs prior to converting the VCF files to Hapmap diploid format
using the TASSEL command-line tools \shortcite{bradbury2007}; variant IDs enabled
retrieval of the proper metadata for each variant after GWAS.

We ran GAPIT using a mixed linear model (MLM) with nine principal components and the VanRaden
algorithm for computing the kinship matrix. We chose to use nine principal components
based on previous work by \shortciteA{deronne2022} using the same population.
Principal components and kinship were computed
on the fly by the GAPIT function itself from the provided inputs, i.e. based on SNPs/indels
or on SVs depending on the analysis. We used an MLM model for
consistency with the statistical model used by the \textit{k}-mer-based
approach.  We used a randomization approach to determine the 5\% family-wise
error-rate threshold for each GWAS analysis as described by \shortciteA{voichek2020}.
To compute this threshold, we permuted the phenotypic observations 100 times
and computed a GWAS on these permuted phenotypes to obtain a distribution of
top \emph{p}-values under the null hypothesis. The fifth most significant of
the top 100 \emph{p}-values was used as the significance threshold for
inference, i.e. markers with \emph{p}-values lower than this threshold were
considered significant.

We conducted GWAS based on \emph{k}-mers using the method developed by
\shortciteA{voichek2020}. In brief, the
analysis was conducted in two steps. The first step used an approximate model
to identify 1 million potentially significant \emph{k}-mers. These were then used in
the second step as input to an exact model implemented in GEMMA \shortcite{zhou2012}, resulting
in a \emph{p}-value for each \emph{k}-mer. We
used the kinship computed from the \emph{k}-mers to correct for relatedness
between accessions. The program automatically outputs a list of \emph{k}-mers
that passed a 5\% family-wise error-rate threshold determined by a
randomization approach, which we considered significant for the purposes of
downstream analyses.

@@

\emph{k}-mers are not intrinsically associated with a particular genomic region
and a tailored analysis was therefore required to associate them with genomic
coordinates. To do so, we queried the mapped reads (in BAM format) of all accessions for matches
to significant \emph{k}-mers for a given trait. Because this is a
computationally demanding operation (there were typically tens of thousands of
significant \emph{k}-mers per trait), we developed a C program called katcher
(\url{https://github.com/malemay/katcher}) that
uses the htslib library \shortcite{bonfield2021} for efficient reading and
writing of BAM files. katcher and its associated utilities allow for efficient
retrieval and annotation of mapped reads containing any \emph{k}-mer from a set
of interest.

Once we retrieved all the reads containing significant \emph{k}-mers, we next
needed to gather the mapping information contained in those reads for the
purposes of identifying contiguous signals and comparing them to annotations. To
simplify the amount of information to process in downstream analyses, we
considered only the most significant \emph{k}-mer contained in a given read and
the position where this \emph{k}-mer was most often observed in
the dataset. Moreover, we limited the analysis to reads with a minimum mapping
quality of 20 and to \emph{k}-mers that were observed at least 10 times in at
least one accession. The latter was done to remove spurious \emph{k}-mers that may originate
from sequencing errors. Filtering based on mapping quality also removed all reads that
mapped to multiple locations because bwa assigns a mapping quality of 0 to such reads.
All downstream analyses (Manhattan plots,
identification of signals, analysis of genes located near significant
\emph{k}-mers, etc.) used this processed set of significant \emph{k}-mers.

@@

\shortciteA{zabala2003} identified Glyma.06g202300 as the gene associated with the
\textit{T} locus for pubescence color and found a 1-bp deletion as a putative
causal variant at this locus. For all three GWAS approaches, we found a signal
overlapping this gene (Supplemental Figure S6). In
particular, the regions defined by the 1\% most associated variants
identified by the SNP- and \textit{k}-mer-based approaches included
Glyma.06g202300 (Supplemental Figure S7). While the most
significant \textit{k}-mers did occur within the sequence of this gene, they
did not correspond to the documented causal variant (Supplemental Figure
S8d).  Still, the fourth and fifth most strongly
associated \textit{k}-mers corresponded to this causal variant (Supplemental Figures
S8d and S9). We
observed that not all accessions with gray pubescence harbored the deleted
nucleotide (Supplemental Figure S9c), which could indicate
either that other causal variants are resulting in a non-functional F3’H in
this population, or that the causal variant lies elsewhere. As for SNP-based GWAS,
the most significant variant within the sequence of Glyma.06g202300 was the
causal variant, however it was not the most significantly associated variant
overall (Supplemental Figures S7b and
S8b). SV-based GWAS did not detect the causal variant
at this locus, as expected from its size.

@@

% Pubescence color - Td locus
We conducted a separate GWAS analysis contrasting only accessions with tawny
and light tawny pubescence because the \textit{Td} locus is known to control
these differences in color. Accordingly, we detected signals in this region
using all three approaches, but only SNP- and \textit{k}-mer-based GWAS
identified a signal that overlapped Glyma.03g258700 (Figure
\ref{pubescence-color-main-figure}a-b, and Supplemental Figures S10 and
S11).  While the most significant
\textit{k}-mer identified the causal variant documented at this locus
\shortcite{yan2020}, it was the fourth most significantly associated variant for
the SNP-based approach (Figure \ref{pubescence-color-main-figure}b-d, Supplemental Figure 
S12).  Interestingly, some of the accessions
bearing the haplotype associated with tawny pubescence displayed light tawny
pubescence, suggesting that there may be more than one causal variant at this
locus (Figure \ref{pubescence-color-main-figure}e).

@@

% Seed coat color - I locus
The causal variant at the \textit{I} locus for seed coat color is a complex SV
that takes the form of a 10.91-kb inverted duplication of three chalcone
synthase (CHS) genes \shortcite<CHS1, CHS3 and CHS4;>[]{tuteja2008}. The presence of
this inverted duplication results in the silencing of all CHS genes in the seed
coat specifically and thus in the absence of seed coat pigmentation
\shortcite{tuteja2009}. While we did find signals overlapping this causal variant
using all three GWAS approaches (Supplemental Figures S13 and
S14), none of the approaches identified
it. This failure to detect the causal variant is likely due
to its complexity, which makes it difficult to detect and genotype it. While it
is likely that at least some of the significantly associated \textit{k}-mers
actually are derived from this causal variant, the lack of systematic methods for
linking \textit{k}-mers to sequence variation hindered our ability to do so.

@@

While all GWAS approaches found signals near the Glyma.19g194300 gene
associated with the \textit{Dt1} locus for stem termination type, the SNP- and
\textit{k}-mer-based analyses performed best since the region defined by their
most highly associated markers (top 1\%) included the causal gene (Supplemental Figures
S18 and S19).
We did identify one of the SNPs suggested as putative causal variants for this
locus by \shortciteA{tian2010} and \shortciteA{liu2010}, however this SNP was only the
135\textsuperscript{th} most associated marker found by Platypus and the
12\textsuperscript{th} most associated \emph{k}-mer (Supplemental Figures
S20 and S21).
Our results are consistent with previous findings in suggesting that there is
more than one causal variant at the \textit{Dt1} locus \shortcite{liu2010,
tian2010}. Indeed, a few cultivars bearing the allele associated with
\emph{dt1} showed indeterminate or semi-determinate phenotypes and some
cultivars showed a determinate phenotype without bearing the allele associated
with \emph{dt1} (Supplemental Figure S21c). The existence of
several causal variants might explain why our analyses did not identify a
causal variant more clearly.

@@

We conducted a GWAS analysis that only compared accessions with black and brown
hila in an attempt to detect stronger signals at the \textit{R} locus.  We
accordingly detected a signal overlapping this locus using all approaches, but
only SNP- and \textit{k}-mer-based GWAS detected signals overlapping the
Glyma.09g235100 gene associated with \textit{R} (Supplemental Figures
S22 and S23).
\shortciteA{gillman2011} identified this R2R3 MYB transcription factor as the
molecular basis for the \textit{R} locus and documented four different
loss-of-function alleles linked to brown hilum color. Although our analyses did
detect strong signals in the vicinity of this gene, the most associated markers
were not located in the body of the gene (Supplemental Figure
S23).  Interestingly, the most significantly
associated \emph{k}-mers detected by our analysis mapped 2 kb upstream of
another gene (Glyma.09g234900) annotated as a MYB transcription factor and
putatively involved in the regulation of anthocyanin biosynthesis. This gene as
well as Glyma.09g235100 and two others are part of a tandem array of four R2R3
MYB genes considered as candidates by \shortciteA{gillman2011}, and of which
Glyma.09g235100 was the only one expressed in the seed coat. Given the
convincing evidence provided by \shortciteA{gillman2011}, it is most likely that
Glyma.09g235100 is indeed the gene associated with the \emph{R} locus and that
our failure to identify the most significantly associated \emph{k}-mers as
causal variants is due to the coexistence of several variants that can cause
loss of function at the \emph{R} locus.  Still, we observed two causal variants
previously documented by \shortciteA{gillman2011} in our dataset (Supplemental Figures
S24 and S25).

@@

We conducted a GWAS analysis comparing accessions with normal and semi-sparse
pubes\-cence in order to detect loci associated with pubescence density. This analysis
detected signals at the \textit{Ps} locus using all three approaches (Figure
\ref{pubescence-density-main-figure}a-b and Supplemental Figures S26, S27
and S28). The causal
variant at this recently cloned locus is a copy number variant (CNV) overlapping the
Glyma.12g187200 gene, with higher copy number resulting in reduced pubescence \shortcite{liu2020ps}.
We identified a region of $\sim$42 kb that varied in copy number across
the population and was associated with the trait (Figure
\ref{pubescence-density-main-figure}c), and therefore possibly corresponded to the
25.6-kb region identified as the causal CNV by \shortciteA{liu2020ps}. We found that
reads containing the most significantly associated \textit{k}-mer mapped to either of the two ends
of the CNV (Figure \ref{pubescence-density-main-figure}d). This suggests that
the most associated \textit{k}-mer is indeed linked to the CNV
that is causal at this locus, while \textit{k}-mers within the
CNV region may represent differences in sequence between the copies.

@@

The difference in size between the CNV reported by \shortciteA{liu2020ps} and our
results may be due to the presence of some unresolved sequence (a stretch of
4,635 N nucleotides) over that interval in the reference sequence that we used.
Because our SV filtering process removed any alternate allele containing N nucleotides,
we reasoned that the CNV may have been found by our pipeline but
filtered out later on. Indeed, we found a 42.8-kb duplication corresponding
to the one described above among the genotype calls made by smoove. An association
analysis of that single variant using GAPIT found a \textit{p}-value of 1.9e-23,
which would have made it the most significant SV. In this particular case, the
failure of the SV-based GWAS analysis to find identify the causal variant was due
to our filtering parameters and not to limitations of the SV genotype calls per se.

@@

We conducted a GWAS analysis comparing dull and shiny seed coat phenotypes for
the seed coat luster trait. Seed coat luster has been shown by
\shortciteA{gijzen1999} to be largely caused by the deposition of a hydrophobic
protein (HPS) at the seed surface. Copy number variation of this gene at the
\textit{B} locus was later shown to explain variation in luster, although
sequence variation in the HPS sequence may also play a role \shortcite{gijzen2006}.
Using a sequencing depth-based analysis, we identified a 31-kb region spanning
positions 9,386,109-9,417,431 on chromosome Gm15 that exhibited variation in
copy number across accessions. This CNV overlapped four genes annotated as
containing HPS domains and therefore likely represented the causal variant
(Supplemental Figures S29 and
S30).  We found signals overlapping this
CNV using all approaches, and the most significantly associated \textit{k}-mer
was notably located within the CNV region (Supplemental Figures
S29 and
S31). However, we were not able to
directly link any of the variants or \textit{k}-mers identified to the causal
CNV.

@@

Both the \textit{L1} and the \textit{L2} loci for pod color have yet to be
cloned. While \shortciteA{he2015} proposed the gene Glyma.19g101700 as likely
corresponding to \textit{L1} based on fine mapping and expression analysis, our
results as well as those reported by \shortciteA{bandillo2017} do not support this
gene as a candidate due to its location $\sim$3 Mb away from GWAS signals. Our
analyses based on \textit{k}-mers and SNPs found signals overlapping 14 and
6 genes, respectively (Supplemental figures S32 and
S33). Of these, a MATE transporter (Glyma.19g120300)
appears as a prime candidate as the most significant \emph{k}-mer was observed
within its sequence and it is highly expressed in pods according to data hosted
on SoyBase \shortcite{grant2010, severin2010}. Moreover, the role of some MATE
proteins in transporting flavonoids has been demonstrated \shortcite{chen2015}.

@@

In addition to \emph{k}-mers, we also assessed the use of SVs in GWAS analysis
of soybean.  Although we did find the causal variant as the most significantly
associated SV at the \emph{W1} locus, results were otherwise rather
inconclusive.  Indeed, most of the other known or putative causal variants were
SNPs or indels, and therefore were not represented in the SV datasets. At most
loci where SVs are known to be the underlying causal variant (e.g.  \emph{I},
\emph{B}, \emph{Ps}), the analyses based on SVs did not find the causal
variants, most likely due to the difficulty of calling and genotyping such
variants \shortcite{kosugi2019, lemay2022}. In the case of the \textit{Ps} locus,
this limitation was due to a filtering step in our pipeline rather than to an
inherent inability to properly call and genotype the causal variant.
